Heuristics and Really Hard Instances for Subgraph Isomorphism Problems

### 2016-01-16 paper/paper.tex
Constrainedness, etc

@@

\begin{abstract}
    We show how to generate ``really hard'' random instances for variants of the subgraph
    isomorphism problem. For the non-induced variant of the problem, we predict and observe a phase
    transition between satisfiable and unsatisfiable instances, with a corresponding complexity peak
    seen in three different solvers. For the induced variant of the problem, the behaviour is more
    complicated, with two phase transitions and an unexpectedly hard region in between. We show that
    this region remains hard when using pseudo-boolean and mixed integer encodings, and under
    reduction to the clique problem. Finally, we show that existing heuristics for the non-induced
    variant can be recovered by determining how to maximise the expected number of solutions, and
    that this technique gives new, better heuristics for the induced variant.
\end{abstract}

@@

            \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
            \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
        \end{scope}
    \end{tikzpicture}\\[0.75cm]
    \begin{tikzpicture}[scale=0.5]
        \begin{scope}
            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgcobalt, inner sep=1.5pt] (N1\n) at (\n, 0) {\n};
            }

            \draw [decorate, decoration={brace, raise=0.2cm}, very thick] (N11.north west) -- (N16.north
                east) node [midway, above=0.3cm] { $1 \mapsto$ };

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgpumpkin, inner sep=1.5pt] (N2\n) at (0, -\n) {\n};
            }

            \draw [decorate, decoration={brace, raise=0.2cm, mirror}, very thick] (N21.north west) --
                (N26.south west) node [midway, left=0.3cm] { $2 \mapsto$ };

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofglawn, inner sep=1.5pt] (N3\n) at (7, -\n) {\n};
            }

            \draw [decorate, decoration={brace, raise=0.2cm}, very thick] (N31.north east) -- (N36.south
                east) node [midway, right=0.3cm] { $3 \mapsto$ };

            \foreach \n in {1, ..., 6}{
                \node [draw, circle, fill=uofgthistle, inner sep=1.5pt] (N4\n) at (\n, -7) {\n};
            }

            \draw [decorate, decoration={brace, raise=0.2cm, mirror}, very thick] (N41.south west) --
                (N46.south east) node [midway, below=0.3cm] { $4 \mapsto$ };

            \begin{scope}[on background layer]
                \draw [color=uofgsandstone!50] ($(N11)!0.8!(N11.south)$) -- ($(N22)!0.8!(N22.east)$);
                \draw [color=uofgsandstone!50] ($(N11)!0.8!(N11.south)$) -- ($(N23)!0.8!(N23.east)$);
                \draw [color=uofgsandstone!50] ($(N11)!0.8!(N11.south)$) -- ($(N24)!0.8!(N24.east)$);
                \draw [color=uofgsandstone!50] ($(N11)!0.8!(N11.south)$) -- ($(N25)!0.8!(N25.east)$);
                \draw [color=uofgsandstone!50] ($(N11)!0.8!(N11.south)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N11)!0.8!(N11.south)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N12)!0.8!(N12.south)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N32)!0.8!(N32.west)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N42)!0.8!(N42.north)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N13)!0.8!(N13.south)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N33)!0.8!(N33.west)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N43)!0.8!(N43.north)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N14)!0.8!(N14.south)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N34)!0.8!(N34.west)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N44)!0.8!(N44.north)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N15)!0.8!(N15.south)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [color=uofgsandstone!50] ($(N21)!0.8!(N21.east)$) -- ($(N45)!0.8!(N45.north)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N22)!0.8!(N22.east)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N42)!0.8!(N42.north)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N23)!0.8!(N23.east)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N43)!0.8!(N43.north)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N24)!0.8!(N24.east)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N44)!0.8!(N44.north)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N25)!0.8!(N25.east)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N45)!0.8!(N45.north)$);
                \draw [color=uofgsandstone!50] ($(N31)!0.8!(N31.west)$) -- ($(N16)!0.8!(N16.south)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N22)!0.8!(N22.east)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N32)!0.8!(N32.west)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N23)!0.8!(N23.east)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N33)!0.8!(N33.west)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N24)!0.8!(N24.east)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N34)!0.8!(N34.west)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N25)!0.8!(N25.east)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [color=uofgsandstone!50] ($(N41)!0.8!(N41.north)$) -- ($(N16)!0.8!(N16.south)$);
                \draw [color=uofgsandstone!50] ($(N12)!0.8!(N12.south)$) -- ($(N33)!0.8!(N33.west)$);
                \draw [color=uofgsandstone!50] ($(N12)!0.8!(N12.south)$) -- ($(N24)!0.8!(N24.east)$);
                \draw [color=uofgsandstone!50] ($(N12)!0.8!(N12.south)$) -- ($(N45)!0.8!(N45.north)$);
                \draw [color=uofgsandstone!50] ($(N22)!0.8!(N22.east)$) -- ($(N14)!0.8!(N14.south)$);
                \draw [color=uofgsandstone!50] ($(N22)!0.8!(N22.east)$) -- ($(N34)!0.8!(N34.west)$);
                \draw [color=uofgsandstone!50] ($(N22)!0.8!(N22.east)$) -- ($(N44)!0.8!(N44.north)$);
                \draw [color=uofgsandstone!50] ($(N22)!0.8!(N22.east)$) -- ($(N16)!0.8!(N16.south)$);
                \draw [color=uofgsandstone!50] ($(N22)!0.8!(N22.east)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N22)!0.8!(N22.east)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N32)!0.8!(N32.west)$) -- ($(N13)!0.8!(N13.south)$);
                \draw [color=uofgsandstone!50] ($(N32)!0.8!(N32.west)$) -- ($(N24)!0.8!(N24.east)$);
                \draw [color=uofgsandstone!50] ($(N32)!0.8!(N32.west)$) -- ($(N44)!0.8!(N44.north)$);
                \draw [color=uofgsandstone!50] ($(N32)!0.8!(N32.west)$) -- ($(N15)!0.8!(N15.south)$);
                \draw [color=uofgsandstone!50] ($(N32)!0.8!(N32.west)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N32)!0.8!(N32.west)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N42)!0.8!(N42.north)$) -- ($(N13)!0.8!(N13.south)$);
                \draw [color=uofgsandstone!50] ($(N42)!0.8!(N42.north)$) -- ($(N24)!0.8!(N24.east)$);
                \draw [color=uofgsandstone!50] ($(N42)!0.8!(N42.north)$) -- ($(N34)!0.8!(N34.west)$);
                \draw [color=uofgsandstone!50] ($(N42)!0.8!(N42.north)$) -- ($(N15)!0.8!(N15.south)$);
                \draw [color=uofgsandstone!50] ($(N42)!0.8!(N42.north)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N42)!0.8!(N42.north)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N13)!0.8!(N13.south)$) -- ($(N34)!0.8!(N34.west)$);
                \draw [color=uofgsandstone!50] ($(N13)!0.8!(N13.south)$) -- ($(N44)!0.8!(N44.north)$);
                \draw [color=uofgsandstone!50] ($(N13)!0.8!(N13.south)$) -- ($(N25)!0.8!(N25.east)$);
                \draw [color=uofgsandstone!50] ($(N13)!0.8!(N13.south)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N23)!0.8!(N23.east)$) -- ($(N15)!0.8!(N15.south)$);
                \draw [color=uofgsandstone!50] ($(N23)!0.8!(N23.east)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [color=uofgsandstone!50] ($(N23)!0.8!(N23.east)$) -- ($(N45)!0.8!(N45.north)$);
                \draw [color=uofgsandstone!50] ($(N23)!0.8!(N23.east)$) -- ($(N16)!0.8!(N16.south)$);
                \draw [color=uofgsandstone!50] ($(N23)!0.8!(N23.east)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N23)!0.8!(N23.east)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N33)!0.8!(N33.west)$) -- ($(N14)!0.8!(N14.south)$);
                \draw [color=uofgsandstone!50] ($(N33)!0.8!(N33.west)$) -- ($(N25)!0.8!(N25.east)$);
                \draw [color=uofgsandstone!50] ($(N33)!0.8!(N33.west)$) -- ($(N45)!0.8!(N45.north)$);
                \draw [color=uofgsandstone!50] ($(N33)!0.8!(N33.west)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N33)!0.8!(N33.west)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N43)!0.8!(N43.north)$) -- ($(N14)!0.8!(N14.south)$);
                \draw [color=uofgsandstone!50] ($(N43)!0.8!(N43.north)$) -- ($(N25)!0.8!(N25.east)$);
                \draw [color=uofgsandstone!50] ($(N43)!0.8!(N43.north)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N14)!0.8!(N14.south)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [color=uofgsandstone!50] ($(N14)!0.8!(N14.south)$) -- ($(N45)!0.8!(N45.north)$);
                \draw [color=uofgsandstone!50] ($(N14)!0.8!(N14.south)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N24)!0.8!(N24.east)$) -- ($(N16)!0.8!(N16.south)$);
                \draw [color=uofgsandstone!50] ($(N24)!0.8!(N24.east)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N24)!0.8!(N24.east)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N34)!0.8!(N34.west)$) -- ($(N15)!0.8!(N15.south)$);
                \draw [color=uofgsandstone!50] ($(N34)!0.8!(N34.west)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N34)!0.8!(N34.west)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N44)!0.8!(N44.north)$) -- ($(N15)!0.8!(N15.south)$);
                \draw [color=uofgsandstone!50] ($(N44)!0.8!(N44.north)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N44)!0.8!(N44.north)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N15)!0.8!(N15.south)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N25)!0.8!(N25.east)$) -- ($(N16)!0.8!(N16.south)$);
                \draw [color=uofgsandstone!50] ($(N25)!0.8!(N25.east)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [color=uofgsandstone!50] ($(N25)!0.8!(N25.east)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N35)!0.8!(N35.west)$) -- ($(N46)!0.8!(N46.north)$);
                \draw [color=uofgsandstone!50] ($(N45)!0.8!(N45.north)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [color=uofgsandstone!50] ($(N45)!0.8!(N45.north)$) -- ($(N36)!0.8!(N36.west)$);
                \draw [ultra thick] ($(N12)!0.8!(N12.south)$) -- ($(N43)!0.8!(N43.north)$);
                \draw [ultra thick] ($(N12)!0.8!(N12.south)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [ultra thick] ($(N12)!0.8!(N12.south)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [ultra thick] ($(N43)!0.8!(N43.north)$) -- ($(N35)!0.8!(N35.west)$);
                \draw [ultra thick] ($(N43)!0.8!(N43.north)$) -- ($(N26)!0.8!(N26.east)$);
                \draw [ultra thick] ($(N35)!0.8!(N35.west)$) -- ($(N26)!0.8!(N26.east)$);
            \end{scope}
        \end{scope}
    \end{tikzpicture}

    \caption{On top, there is a non-induced isomorphism from the pattern graph, in the center, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6, 3 to
    5 and 4 to 3, is both non-induced and induced.
    Below, the association graph encoding for the induced version: the highlighted clique
    corresponds to the same mapping.}
    \label{figure:sip}
\end{figure}

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices.  This is not
ideal: such instances are guaranteed to be satisfiable (?? mention \citep{Anton:2009}), and so
existing benchmark suites contain relatively few non-trivial unsatisfiable instances. These
instances also tend to be computationally fairly easy, with most of the difficulty being in dealing
with the size of the model.

Generating non-trivial unsatisfiable random instances is not straightforward. Using a
pattern graph from one of the random suites with the ``wrong'' target graph tends to give either a
trivially unsatisfiable instance, or a satisfiable instance. (In particular, it is \emph{not} the
case that a relatively small random graph is unlikely to appear in a larger random graph.)

@@

?? Maybe introduce \cref{figure:phase-transition} here instead of at the start of the next section?

For subgraph isomorphism we have three relevant control parameters: we can independently alter the
edge probability of the pattern graph, the edge probability of the target graph, and the relative
orders (number of vertices) of the pattern and target graphs.  For non-induced isomorphisms, with
the correct choice of parameters we see results very similar to those observed with boolean
satisfiability problems: there is a phase transition from satisfiable to unsatisfiable (and we can
predict its location), and we see a complexity peak occur (with three different solvers) near this
phase transition.

?? Actually, it's not necessarily obvious that these are the right parameters to use... For example,
mean degree is a meaningful parameter for random graphs... Also, the formula shows it's not exactly
the ``relative'' order, but some other function of the two orders.

For certain choices of parameters for induced isomorphisms, there are two phase transitions, going
from satisfiable to unsatisfiable, and then from unsatisfiable back to satisfiable. Again, when
going from satisfiable to unsatisfiable (from either direction), instances go from being trivial to
really hard to solve. However, each of the three solvers we tested also finds the central
unsatisfiable region to be really hard---this may be unexpected, since this region is not near a
phase transition and is apparently over-constrained. To show that this is not a simple weakness of
current subgraph isomorphism algorithms, we verify that this region is also hard when using use a
pseudo-boolean encoding, a mixed integer model, and under reduction to the clique problem.

?? Also there's heuristics. Figure out how to work this smoothly into the intro.

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops (vertices which are
adjacent to themselves).  The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\overline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ iff
$v$ and $w$ are not adjacent in $G$. We write $G(v, p)$ for an Erd\H{o}s-R\'enyi random graph with
$v$ vertices, and an edge between each distinct pair of vertices with independent probability $p$.

?? I'm sure I figured out how to make overlines typeset better. Remember to look this up.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. An \emph{induced subgraph
isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$,
then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P \rightarrowtail T$
for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced isomorphism. Observe
that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced isomorphism $i : P
\rightarrowtail T$ which is also a non-induced isomorphism $\overline{i} : \overline{P}
\rightarrowtail \overline{T}$.

The \emph{association graph encoding} of a subgraph isomorphism problem is constructed by creating a
new graph with a vertex for each pair $(p, t)$ of vertices from the pattern and target graphs
respectively. There is an edge between vertex $(p_1, t_1)$ and vertex $(p_2, t_2)$ if mapping $p_1$
to $t_1$ and $p_2$ to $t_2$ simultaneously is permitted. For the non-induced problem, permitted
means either $p_1$ is not adjacent to $p_2$, or both $p_1$ and $p_2$, and $t_1$ and $t_2$, are
adjacent; for the induced problem, it means that $p_1$ is adjacent to $p_2$ if and only if $t_1$ is
adjacent to $t_2$. Either way, it can be seen that a clique (a subset of vertices, each of which is
adjacent to every other in the subset) of size equal to the order of the pattern graph exists in the
association graph if and only if the problem is satisfiable, and the vertices selected correspond to
the assignments making the solution \citep{Levi:1973}. We illustrate this in
\cref{figure:sip}. ?? Can we move this to section 4? I thought we'd need this for the sol formulae,
but it appears not.

@@

?? Stuff about runtimes and recursive calls. We are not aiming to compare solvers; rather, we are
looking for solver-independent behaviour.

@@

\begin{figure}[tb]
    \input{gen-graph-phase-transition.tex}
    \caption{With a fixed a pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.4, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort, and mean proportion satisfiable, using a
    larger sample size.}
    \label{figure:phase-transition}
\end{figure}

@@

\begin{figure}[tb]
    \input{gen-graph-non-induced.tex}
    \caption{Behaviour of algorithms on the non-induced variant. Each point is the average of ten
        runs. For each plot, the x-axis is the pattern edge probability and the y-axis is the target
        edge probability, both from 0 to 1. Along the top row, we show the proportion of instances which are
        satisfiable; the white bands shows the phase transitions, and the black lines are our
        predictions of where the phase transition will occur. On the second row, we show the number
    of search nodes used by the Glasgow algorithm, on the third row, the number of search nodes used
by the LAD algorithm, and the fourth, VF2; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

@@

On subsequent rows, we show the average number of search nodes used by the different algorithms. In
general, satisfiable instances are easy, until very close to the phase transition. As we hit the
phase transition and move into the unsatisfiable region, we see a complexity increase. Finally, as
we pass the phase transition and move deeper into the unsatisfiable region, instances become easier
again. This behaviour is largely solver-independent, although VF2 has a larger hard region than
Glasgow or LAD. Thus, although we have moved away from the single control parameter typically
studied, ?? I guess we want to say it's not that surprising?

@@

?? Note that the size of the state space is defined differently to in binary CSPs: we're using the
all diff constraint to determine the size of the state space, and are not treating it as a
constraint. In other words, our generate-and-test approach is alldiff-aware. There are probably five
people in the world who care about this and would have seen it as very interesting twenty years ago.

@@

?? Point out how the value ordering heuristics, which were developed empirically, can be recovered
just by using this formula.  Note that the Glasgow heuristics are moving us towards the satisfiable
region, or increasing the expected number of solutions \citep{Gent:1996:EN}. LAD does SDF with no
tiebreaking, no value ordering heuristic.

?? In contrast, VF2 heuristics work by selecting vertices adjacent to vertices already selected.

?? I suppose we should mention kappa \citep{Gent:1996:Kappa}, etc, here (or possibly in the
conclusion?). This whole thing is a bit subtle: maximising E(N) or Kappa as a principle is
encoding-independent, but whether this is done by SDF, deg, etc, is encoding-dependent. This seems
to be known by these same five people, but maybe isn't in the literature anywhere.

@@

\begin{figure*}[tb]
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second, third and fourth rows show the number of search nodes used by the
    Glasgow, LAD and VF2 algorithms. The fifth row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs. The final row shows when the Glasgow algorithm performs better when given the
    complements of the pattern and target graphs as inputs.}\label{figure:induced}
\end{figure*}

@@

@@

?? Does kappa predict the hard region in the middle?

>>>>

% cref style
\crefname{figure}{Figure}{Figures}
\Crefname{figure}{Figure}{Figures}

% http://tex.stackexchange.com/questions/22100/the-bar-and-overline-commands
\newcommand{\shortoverline}[1]{\mkern 1.5mu\overline{\mkern-1.5mu#1\mkern-1.5mu}\mkern 1.5mu}

@@

\begin{abstract}
    We show how to generate ``really hard'' random instances for variants of the subgraph
    isomorphism problem. For the non-induced variant, we predict and observe a phase
    transition between satisfiable and unsatisfiable instances, with a corresponding complexity peak
    in three different solvers. For the induced variant, much richer behaviour is observed, and we
    see evidence that constrainedness is a better measure of difficulty than proximity to a phase
    transition. Finally, we show that existing heuristics for the non-induced variant can be
    recovered by determining how to maximise the expected number of solutions, and that this
    technique gives new, better heuristics for the induced variant.
\end{abstract}

@@

            \node [anchor=center, font=\Large] (A1) at (-1.5, -1.5) { $\hookleftarrow$ };
            \node [anchor=center, font=\Large] (A2) at ( 3.5, -1.5) { $\rightarrowtail$ };
        \end{scope}
    \end{tikzpicture}

    \caption{There is a non-induced isomorphism from the pattern graph, in the center, to the target
    on the right, mapping vertex 1 to 1, 2 to 3, 3 to 5 and 4 to 6. This is not an induced
    isomorphism, since there is an edge between 1 and 5 in the target but not between 1 and 3 in the
    pattern. The mapping from the pattern to the (same) target on the left, sending 1 to 2, 2 to 6,
    3 to 5 and 4 to 3, is both non-induced and induced.}
    \label{figure:sip}
\end{figure}

However, these algorithms cannot handle \emph{arbitrary} instances of this size. The experimental
evaluations of these algorithms were performed using a mix of real-world graphs, graphs that encode
biochemistry and computer vision problems, and randomly generated graph pairs. Using random
instances to evaluate algorithm behaviour can be beneficial, because it provides a way of generating
many instances cheaply, and reduces the risk of over-fitting when tuning design parameters. The
random instances used in each case came from common datasets \citep{DeSanto:2003,Zampelli:2010},
which were generated by taking a random subgraph of a random graph (using various models, including
Erd\H{o}s-R\'enyi, scale-free, bounded degree, and meshes) and permuting the vertices. Such
instances are guaranteed to be satisfiable---\citet{Anton:2009} exploited this property to create
large sets of random satisfiable SAT instances.  However, since this has been the only approach used
to generate random subgraph isomorphism instances, existing benchmark suites contain relatively few
non-trivial unsatisfiable instances, and the satisfiable instances tend to be computationally fairly
easy, with most of the difficulty being in dealing with the size of the model.  This problem cannot
be addressed by using a pattern graph from one of the random suites with the ``wrong'' target graph:
this tends to give either a trivially unsatisfiable instance, or a satisfiable instance. (In
particular, it is \emph{not} the case that a relatively small random graph is unlikely to appear in
a larger random graph.)

@@

For subgraph isomorphism we identify relevant control parameters: we can independently alter the
edge probability of the pattern graph, the edge probability of the target graph, and the relative
orders (number of vertices) of the pattern and target graphs.  For non-induced isomorphisms, with
the correct choice of parameters we see results very similar to those observed with boolean
satisfiability problems: there is a phase transition from satisfiable to unsatisfiable (and we can
predict its location), and we see a complexity peak occur with three different solvers near this
phase transition.

For certain choices of parameters for induced isomorphisms, there are two phase transitions, going
from satisfiable to unsatisfiable, and then from unsatisfiable back to satisfiable. Again, when
going from satisfiable to unsatisfiable (from either direction), instances go from being trivial to
really hard to solve. However, each of the three solvers we tested also finds the central
unsatisfiable region to be really hard---this may be unexpected, since this region is not near a
phase transition. To show that this is not a simple weakness of current subgraph isomorphism
algorithms, we verify that this region is also hard when using use a pseudo-boolean encoding, and
under reduction to the clique problem. Interestingly, the constrainedness measure
\citep{Gent:1996:Kappa} \emph{does} predict this difficult region---these instances provide (?? the
first??) evidence in favour of constrainedness, rather than proximity to a phase transition, being
an accurate predictor of difficulty.

?? Finally, we discuss the relationship between heuristics and the expected number of solutions,
recovering existing heuristics for the non-induced variant, and introducing new heuristics for the
induced variant. I have failed miserably at integrating this paragraph smoothly.

@@

Throughout, our graphs are unlabelled, undirected, and do not have any loops (vertices which are
adjacent to themselves).  The \emph{order} of a graph is the cardinality of its vertex set. We write
$\operatorname{V}(G)$ for the vertex set of a graph $G$. The \emph{complement} of a graph $G$,
denoted $\shortoverline{G}$, is the
graph with the same vertex set as $G$, and with an edge between distinct vertices $v$ and $w$ iff
$v$ and $w$ are not adjacent in $G$. We write $G(v, p)$ for an Erd\H{o}s-R\'enyi random graph with
$v$ vertices, and an edge between each distinct pair of vertices with independent probability $p$.

A \emph{non-induced subgraph isomorphism} from a graph $P$ (called the \emph{pattern}) to a graph
$T$ (called the \emph{target}) is an injective mapping from $\operatorname{V}(P)$ to
$\operatorname{V}(T)$ which preserves adjacency---that is, for every adjacent $v$ and $w$ in
$\operatorname{V}(P)$, the vertices $i(v)$ and $i(w)$ are adjacent in $T$. An \emph{induced subgraph
isomorphism} additionally preserves non-adjacency---that is, if $v$ and $w$ are not adjacent in $P$,
then $i(v)$ and $i(w)$ must not be adjacent in $T$. We use the notation $i : P \rightarrowtail T$
for a non-induced isomorphism, and $i : P \hookrightarrow T$ for an induced isomorphism. Observe
that an induced isomorphism $i : P \hookrightarrow T$ is a non-induced isomorphism $i : P
\rightarrowtail T$ which is also a non-induced isomorphism $\shortoverline{i} : \shortoverline{P}
\rightarrowtail \shortoverline{T}$.

@@

?? Stuff about runtimes and recursive calls. We are not aiming to compare solvers; rather, we are
looking for solver-independent behaviour. We need a quick note on the models used here too.

@@

\begin{figure}[tb]
    \setlength{\abovecaptionskip}{6pt}
    \input{gen-graph-phase-transition.tex}
    \caption{With a fixed a pattern graph order of 20, a target graph order of 150, a target edge
        probability of 0.4, and varying pattern edge probability, we observe a phase transition and
        complexity peak with the Glasgow solver in the non-induced variant. Each point represents
        one instance. The lines show mean search effort, and mean proportion satisfiable.}
    \label{figure:phase-transition}
\end{figure}

@@

\begin{figure}[tb]
    \setlength{\abovecaptionskip}{0pt}
    \input{gen-graph-non-induced.tex}
    \caption{Behaviour of algorithms on the non-induced variant. Each point is the average of ten
        runs. For each plot, the x-axis is the pattern edge probability and the y-axis is the target
        edge probability, both from 0 to 1. Along the top row, we show the proportion of instances which are
        satisfiable; the white bands shows the phase transitions, and the black lines are our
        predictions of where the phase transition will occur. On the second row, we show the number
    of search nodes used by the Glasgow algorithm, on the third row, the number of search nodes used
by the LAD algorithm, and the fourth, VF2; the dark regions indicate ``really hard'' instances.}
    \label{figure:non-induced}
\end{figure}

@@

On subsequent rows, we show the average number of search nodes used by the different algorithms. In
general, satisfiable instances are easy, until very close to the phase transition. As we hit the
phase transition and move into the unsatisfiable region, we see a complexity increase. Finally, as
we pass the phase transition and move deeper into the unsatisfiable region, instances become easier
again. This behaviour is largely solver-independent, although VF2 has a larger hard region than
Glasgow or LAD. Thus, although we have moved away from the single control parameter typically
studied, we observe similar behaviour to other \NP-complete problems.

@@

?? Note that the size of the state space is defined differently to in binary CSPs: we're using the
all diff constraint to determine the size of the state space, and are not treating it as a
constraint. In other words, our generate-and-test approach is alldiff-aware. This matters for kappa.

@@

When designing variable and value ordering heuristics for backtracking search algorithms, there are
various general principles which appear to be effective---one of these is to try to maximise the
expected number of solutions inside any subproblem considered during search \citep{Gent:1996:EN}.
This is usually done by cheaper surrogates, rather than direct calculation. When branching, both LAD
and Glasgow will pick a variable with fewest remaining values in its domain: doing this will
generally reduce the first part of the $\langle Sol \rangle$ equation by as little as possible. When
two or domains are of equal size, LAD simply breaks ties lexicographically. Glasgow, however, will
pick a variable corresponding to a pattern vertex of highest degree. This strategy was determined
empirically, but could have been derived from the $\langle Sol \rangle$ formula: picking a
pattern vertex of high degree will make the remaining pattern subgraph sparser, which will decrease
the exponent in the second half of the formula, maximising the overall value. LAD does not apply a
value ordering heuristic, but Glasgow does: it prefers target vertices of lowest degree. Again, this
was determined empirically, but it has the effect of increasing $\langle Sol \rangle$ by increasing
the remaining target density.

The VF2 heuristics, in contrast, are based around preserving connectivity. This perhaps explains
some of VF2's weakness on graphs that are not extremely sparse.

@@

\begin{figure*}[tb]
    \setlength{\abovecaptionskip}{0pt}
    \input{gen-graph-induced.tex}
    \caption{Behaviour of algorithms on the induced variant, shown in the style
    of \cref{figure:non-induced}. The second, third and fourth rows show the number of search nodes used by the
    Glasgow, LAD and VF2 algorithms. The fifth row shows a bound on the satisfiable region, by
    considering where a \emph{non-}induced isomorphism may also be a non-induced isomorphism between
    complement graphs. The final row shows when the Glasgow algorithm performs better when given the
    complements of the pattern and target graphs as inputs. ?? Also plot constrainedness
here.}\label{figure:induced}
\end{figure*}

@@

\subsection{Why is the Central Region Hard?}

@@

?? Constrainedness predicts the hard region in the middle \citep{Gent:1996:Kappa}! Add a two line
description of kappa, the slightly modified way we calculated it, and plot it on the graph.
